---
output: md_document
---
# PA1 for the reproducible research class

## Introduction

This Readme file performs the analysis required in order to complete the 
Project assignement 1 for the reproducible research class. 

In order to be successfully reviewed, we need to allow for our r code chunks
to be seen, we therefore set the global option

```{r setoption}
library(knitr)
opts_chunk$set(echo=TRUE)
```

## Loading the data

The first step towards the completion of the project is to load the data

```{r}
activity<-read.csv("activity.csv",sep=",",header=TRUE)
```

## Number of steps per day, mean and median

We use the very handy tapply function, the index being the date, we store
the result in a variable `sumstep` for future analysis.**Note that
we exclude the days where there are only NA values from our analysis. 
Consequently, these days (there are 8 of them) don't appear in sumstep** 

```{r}
sumstep<-tapply(activity$steps, activity$date,  FUN =sum)[
    !is.na(tapply(activity$steps, activity$date,  FUN =sum))]
sumstep
```

We then display the histogram  of th total number of steps per day
```{r,fig.height=6}
hist(sumstep,breaks=5, col="cornflowerblue", border="white",
    main="histogram of the total number of steps per day",
    xlab=" total number of steps per day", ylab="Number of occurences",
    xlim=c(0,25000))
```

Finally, the mean the median are just computed using the corresponding 
functions, thanks to the `sumstep` variable previously stored

```{r}
mean(sumstep)
median(sumstep)
```

## averaged number of steps per interval

We use again the tapply function, the index being now the interval.
We store the result in a variable`meaninterval` Note that we exclude
the NA values from our analysis (we don't show the result to save place)
```{r}
meaninterval<-tapply(activity$steps, activity$interval,  FUN =mean,na.rm=TRUE)
```

We know make the required time series plot

```{r,fig.height=6}
plot(activity$interval[1:length(meaninterval)],meaninterval,type="l",
     xlab="5mn interval",ylab="averaged number of steps over all days",
     col="cornflowerblue",lwd=3,
     main="averaged number of steps over all days per 5 mn interval")
```

The maximum 5 mn interval is easily found thanks to 
```{r}
names(which.max(meaninterval))
```

which corresponds to the value
```{r}
meaninterval[which.max(meaninterval)]
```

## Missing data

To find the number of missing data, we just use apply on the first column of
our dataframe

```{r}
sum(apply(activity,1,FUN=is.na))
```

To fill these missing data, we will use one of the strategies advocated by the 
project assignement, namely replacing the NA by the correponding 5 mn averaged 
number of steps over all days


```{r}
a<-replace(activity$steps,is.na(activity$steps),meaninterval)
```

we then create our new dataset without NA values
```{r}
fullactivity<-activity
fullactivity$steps<-a
```

Defining the variable corresponding to the total number of steps taken each day

```{r}
fullsumstep<-tapply(fullactivity$steps, fullactivity$date,  FUN =sum)
fullsumstep
```

we do the same histogram as in part 1 with this new dataset

```{r,fig.height=6}
hist(fullsumstep,breaks=5, col="cornflowerblue", border="white",
    main="histogram of the total number of steps per day",
    xlab=" total number of steps per day", ylab="Number of occurences",
    xlim=c(0,25000))
```

Finally, the mean the median are just computed using the corresponding 
functions, thanks to the ```fullsumstep``` variable previously stored

```{r}
mean(fullsumstep)
median(fullsumstep)
```

we find that the mean is unchanged, while the median is now slightly changed,
and equal to the mean. This makes perfect sense, as we completed the missing
data (corresponding to 8 full days) with the 5 mn averaged over all days, and
summing over the full day it gives a total number of steps equal to the mean. 
So the mean is obviously left unchanged, and since we added 8 "mean values"
the median is now equal one of these values, hence the mean.

## Impact of the day of the week

TO create our new factor we start by creating our weekday vector (sorry for
the french names :)
```{r}
weekday<-c("lundi","mardi","mercredi","jeudi","vendredi")
```

and we then create the new column of our dataframe with the ifelse command
```{r}
fullactivity$weektime<-as.factor(ifelse(weekdays(
    strptime( fullactivity$date,"%Y-%m-%d")) %in% weekday,"weekday","weekend"))
```

With this additional index, we create the 5 mn interval averaged over either
all weekdays or all weekends

```{r}
weekmeaninterval<-tapply(fullactivity$steps,
    list(fullactivity$interval,fullactivity$weektime),  FUN =mean)
```

That allow us to perform the required analysis

```{r,fig.height=8,fig.width=10}
 par(mfrow=c(1,2))
plot(fullactivity$interval[1:length(weekmeaninterval[,1])],weekmeaninterval[,1],type="l",
    xlab="5mn interval",ylab="averaged number of steps over all days",
    col="cornflowerblue",lwd=3,
    main="weekdays")
plot(fullactivity$interval[1:length(weekmeaninterval[,2])],weekmeaninterval[,2],type="l",
    xlab="5mn interval",ylab="averaged number of steps over all days",
    col="cornflowerblue",lwd=3,
    main="weekends")
```